An Efficient Parameter Space Search as an Alternative to Markov Chain Monte Carlo

### 2014-07-21 aps4p5/drafts/draft_140719.tex
latest draft of paper

We consider the following problem.  A researcher wants to constrain a theory
described by some number $N_p$ of tunable parameters, which we will represent by
the $N_p$-dimensional vector $\vec{\theta}$.  The available data measure some
function $f$ which is an implicit function of $\vec{\theta}$.  The researcher
is able to calculate some statistic (e.g. $\chi^2$) quantifying the fit of a 
model $f(\vec{\theta})$ to the data and wishes to find a confidence limit
defined by some rule, e.g $\chi^2\le\chi^2_\text{lim}$.  This is a general
phrasing of a Frequentist confidence limit as discussed in Appendix
\ref{sec:frequentism}.  We discuss alternatives for determining
$\chi^2_\text{lim}$ in Section \ref{sec:chi} below.

@@

Markov Chain Monte Carlo (MCMC) methods propose to find the desired
limits by drawing random samples from the parameter space which are drawn
according to a probability density (the Bayesian posterior) which can be
interpreted as the probability that some value of $\vec{\theta}$ represents the
true value of $\vec{\theta}$ realized by the physical process underlying
the data.  
With enough of these samples, one can integrate this
probability density and find regions which contain some set amount of the total
probability.  In this case, ``credible limits'' are phrased as target values
for the probability integral, i.e. 
$0.68$ for ``1-$\sigma$'' limits, $0.95$ for ``2-$\sigma$'' limits
and so on \cite{mcmc}.

We propose an alternative solution, using methods from machine learning
(to wit, Gaussian processes, though the algorithm admits many possible drivers; see
Section \ref{sec:gp})
to find the desired confidence limit by directly searching the
parameter space of $\vec{\theta}$ for points which lie on the boundary of the
desired confidence limit $\chi^2\le\chi^2_\text{lim}$.  
We refer to this algorithm as Active Parameter Searching (APS).
We present the algorithm in detail in Sections \ref{sec:algorithm} and
\ref{sec:gp}.
As it searches, APS uses
the knowledge it has already learned about $\vec{\theta}\rightarrow\chi^2$ 
to improve its search according to a metric that rewards the
algorithm both for finding points on the confidence limit and for sampling
points from previously unexplored regions of parameter space.  We find that this
behavior makes APS more robust against multi-modal $\chi^2$ functions (see
Section \ref{sec:toy}) without
sacrificing the speed of convergence 
associated with traditional MCMC methods (see Section \ref{sec:wmap}).
We show that APS is capable of yielding both Frequentist confidence
limits (which we define in Appendix \ref{sec:frequentism}) as well as more traditional
Bayesian credible limits on the space of $\vec{\theta}$.

@@

\begin{itemize}
\item(1A) Generate some initial number $N_s$ of randomly-distributed points in 
parameter space.  Find the values of $\chi^2$ corresponding to these points.
\\
\item(2A) Generate some number $N_c$ of candidate points.  
For each of these points, find the $N_g$ nearest points already sampled
by APS. Use these nearest neighbor points and their corresponding $\chi^2$ values
to predict the value of $\chi^2$ at each of these points.
 We will signify this prediction by $\mu$.  Calculate some value $\sigma$
 quantifying confidence in $\mu$ as a prediction of $\chi^2$.  Low values of
 $\sigma$ should correspond to confident predictions.
 \\
\item(3A) Choose the candidate point which maximizes the statistic
\begin{equation}
\label{eqn:sstat}
S=\sigma-|\mu-\chi^2_\text{lim}|
\end{equation}
and evaluate $\vec{\theta}\rightarrow\chi^2$ at that point.  Add this point and
its $\chi^2$ value to the list of sampled points begun in (1A). 
Return to (2A) and repeat until convergence.
\\
\end{itemize}
The confidence limit reported by APS is the list of all points found
for which $\chi^2\le\chi^2_\text{lim}$.  
These points, when plotted in one- or two-dimensional slices of parameter
space, ought to sketch out a region which contains the true value of $\vec{\theta}$
with the desired confidence (as defined in Appendix \ref{sec:frequentism}).
This is in contrast to the credible
limits of MCMC, which represent an integral over the sampled points containing some
fraction of the total posterior probability.

@@

The first possibility, a user-defined $\chi^2_\text{lim}$ is based
on the theory of Frequentist confidence intervals, which we briefly discuss
here and consider in much greater detail in Appendix \ref{sec:frequentism}.
If a data set is comprised of $N_d$ Gaussian-distributed data points
with a known covariance structure, then 
\begin{equation}
\chi^2\equiv\sum_{i,j}^{N_d}(d_i-f_i(\vec{\theta}))C^{-1}_{i,j}
(d_j-f_j(\vec{\theta}))
\end{equation}
will be distributed according to a $\chi^2$-distribution with $N_d$ degrees
of freedom.  This distribution can be integrated to determine the value
of $\chi^2$ containing $(1-\alpha)\%$ of the total probability.  This limiting
value will be the user-defined $\chi^2_\text{lim}$.

If the user does not feel comfortable settig $\chi^2_\text{lim}$
a priori, APS $\chi^2_\text{lim}$ can learn $\chi^2_\text{lim}$ by
defining it as $\chi^2_\text{min}+\Delta\chi^2$.  Here, $\chi^2_\text{min}$
is the minimum value of $\chi^2$ discovered by APS.  
$\Delta \chi^2$ is set
based on the properties of the $N_p$-dimensional parameter space being explored.
To wit, Wilks (1938) proves that $\Delta\chi^2$ for an $N_p$-dimensional
parameter space will be distributed according to a $\chi^2$-distribution
with $N_p$ degrees of freedom.  One can perform the same integral discussed
above and set $\Delta\chi^2$ to that value which encloses the desired confidence
limit.  This is the well-known Likelihood Ratio Test \cite{np}.
In this case, the user is relying on APS to eventually find the true
$\chi^2_\text{min}$ of the likelihood surface being explored.  
Section \ref{sec:simplex} below outlines an additional means of search
by which our APS is capable of efficiently locating $\chi^2_\text{lim}$.

@@

Bryan {\it et al.} (2007) show that the APS search algorithm as presented in steps (1A-3A)
is robust against multi-modal
likelihood functions, using it to identify a disjoint region of high likelihood in
cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
spectrum.  This is the principal advantage to selecting points for evaluation according
to the statistic (\ref{eqn:sstat}).  Unfortunately, it is possible that the focus on 
exploring
unknown regions of parameter space might slow down convergence in the case of uni-modal
likelihood functions.  Therefore, we propose the following modified search
to ensure that it adequately characterizes known regions of high likelihood while
simultaneously searching for new regions to explore.

These modified searches will require APS to keep track of information it learns about the
$\vec{\theta}\rightarrow\chi^2$ fuction as it searches.  
To facilitate this discussion, we introduce the folling notations.
Symbols denoted with $\vec{x}$ represent points in parameter space.  Variables denoted with
$\{\hat{X}\}$ denote sets of points in parameter space.  Variables denoted $\{X\}$ are sets
of numbers stored by APS.

@@

For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, APS performs the following:
\begin{itemize}
\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, 
randomly choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
evaluate $\chi^2$ at that point.  
\\
\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
use bisection
along the line connecting the proposed point to $\vec{M}$ to find a point which does
satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way \
with $\chi^2=\chi^2_\text{lim}$ are stored in the set $\{\hat{B}(\vec{M})\}$ of 
``boundary points'' about $\vec{M}$ (each $\vec{M}$ in $\{\hat{\Gamma}\}$ will have its own
set $\{\hat{B}(\vec{M})\}$).  
\\
\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
Instead, examine the points in $\{\hat{B}(\vec{M})\}$ and identify the points
corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
points will be referred to as $\{\hat{B}_\text{MaxMin}(\vec{M})\}$.  
There will be $2\times N_p$ points in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, one for
the maximum and one for the minimum in each of our $N_p$ parameters.  
\\
\item(4B) For each point $\vec{b}$ in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, 
propose some number $N_v$ of new points (we have chosen
$N_v=20$ for the sake of speed) $\vec{v} = \vec{b}+\vec{\epsilon}$ where $\vec{\epsilon}$ is a
random, small vector in parameter space constrained to be perpendicular to
$\vec{b}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{v}$
($N_v$ such points for each of the $2\times N_p$ extremal boundary points in
$\{\hat{B}_\text{MaxMin}(\vec{M})\}$).
\\
\item(5B) At each point $\vec{v}$ above, evaluate equation (\ref{eqn:sstat}).  
Evaluate $\chi^2$ at the point $\vec{v}$ which maximizes $S$. If
$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
\end{itemize}

To reiterate, our version of APS alternates between performing steps (1A-3A) and steps
(1B-5B).  If there are $N_m$ distinct known local minima in $\{\hat{\Gamma}\}$, 
our algoritm will perform
$N_m$ iterations of steps (1A-3A), and then perform steps (1B-5B) once for each
$\vec{M}$ in $\{\hat{\Gamma}\}$.

@@

The modification suggested in section \ref{sec:focus} is designed to explore 
$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
perpendicularly to its ``radii'' (the vector $\vec{b}-\vec{M}$ would be a radius of
$\chi^2=\chi^2_\text{lim}$ described a sphere).  It is, however, also possible
that the $S$-maximization in steps (1A-3A) could also learn something about the
shape of the $\chi^2=\chi^2_\text{lim}$ contour, even if it does not place
points directly on that contour.  To explore that possibility, we add the
following steps to the end of the sequence (1A-3A).

\begin{itemize}
\item(1C) For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, store, 
in addition to the boundary
points $\{\hat{B}(\vec{M})\}$, the projection of those boundary points
onto a unit sphere in parameter space, i.e. for each point $\vec{b}$ in $\{\hat{B}(\vec{M})\}$, draw a line
connecting $\vec{b}$ with $\vec{M}$ and keep track of
where a unit sphere centered on $\vec{M}$ intersects that line.  Call this set
of projectsion $\{\hat{U}\}$.
\\
\item(2C) Call the point for which $\chi^2$ has been evaluated in step (3A)
$\vec{P}$., find the
point $\vec{M}$ in $\{\hat{\Gamma}\}$ that is nearest to $\vec{P}$ in parameter space.
\\
\item(3C) Find the
projection of $\vec{P}$ onto the unit sphere surrounding $\vec{M}$.  
Call this projected point $\vec{p}$.  Find the point $\vec{u}$ in
$\{\hat{U}\}$ that is nearest to $\vec{p}$.  Record the parameter space 
distance between $\vec{u}$ and $\vec{p}$.
Add this distance to the set $\{D\}$ ($\{D\}$ will be a running list
of the $\vec{u}-\vec{p}$ distance for all points $\vec{P}$ found by steps
1A-3A, regardless of which point in $\vec{M}$ they are nearest).
\\
\item(4C) If the $\vec{p}-\vec{u}$ distance is greater than the 2/3 quantile of $\{D\}$,
then perform bisection between $\vec{M}$ and the point evaluated in step (3A).
\end{itemize}

This way, points on the $\chi^2=\chi^2_\text{lim}$ contour can be found whose
directions away from $\vec{M}$ in parameter space have not yet been explored.
Because we expect the unit spheres surrounding the points $\vec{M}$ to become more densely
populated as the algorithm runs, we periodically delete the list $\{D\}$ so that
new directions away from $\vec{M}$ can still be explored.

@@

The final modification to the original APS algorithm which we propose in this
work is to utilize function minimization to identify local minima in $\chi^2$.
One of the principal drawbacks of the algorithm presented in steps (1A-3A) is
that it makes no account of ``near misses.''  If step (3A) finds that $\chi^2$
is lower than average but still greater than $\chi^2_\text{lim}$, the original
APS algorithm does nothing with that information.  We have tried to improve upon
that with our incorporation of bisection in sections \ref{sec:focus} and
\ref{sec:unitSphere} above.  Our final innovation is to allow APS to use points
with 
\begin{eqnarray}
\chi^2_\text{lim}&\le&\chi^2\le\chi^2_\text{threshold}\label{eqn:simplexCriterion}\\
\chi_\text{threshold}&=&\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})
\label{eqn:threshold}
\end{eqnarray}
to seed a simplex function minimizer as developed by Nelder and Mead (1965).
Here, $G$ is a user-set parameter, $\chi^2_\text{min}$ is the minimum discovered
$\chi^2$ value and $\chi^2_{1/10}$ is the 1/10 quantile of the $\chi^2$ values
discovered by steps (1A-3A) (not those discovered by any of the modifications
discussed in sections \ref{sec:focus}, \ref{sec:unitSphere} or
\ref{sec:simplex}).

Because the Nelder-Mead simplex minimizer can be very time intensive, we balance our
algorithm by demanding that the number of $\chi^2$ evaluations devoted to all
other variants of search (i.e. steps (1A-3A) and (1B-5B)) equal the number of
$\chi^2$ evaluations devoted to the simplex minimizer before a new simplex
minimization can occur.

If the simplex minimizer converges to a point such that
$\chi^2<\chi^2_\text{lim}$, our APS determines whether or not it
has found a new local minimum to be added to $\{\hat{\Gamma}\}$ by comparing the end
point $\vec{y}$ of the minimization to each of the points already in
$\{\hat{\Gamma}\}$.  For each $\vec{M}$ in $\{\hat{\Gamma}\}$, APS calculates the mid-point
\begin{equation}
\label{eqn:midpt}
\vec{m}=0.5\times\left(\vec{y}+\vec{M}\right)
\end{equation}
and $\vec{M}$.  If, for all $\vec{M}$, 
$\chi^2$ at $\vec{m}$ is greater
than $\chi^2_\text{lim}$, then it is deemed that $\vec{\gamma}$ is a new local
minimum and is added to $\{\hat{\Gamma}\}$.  If not, $\vec{\gamma}$ is assumed to be a
part of the one of the previously identified loci of low $\chi^2$ and it is not
added to $\{\hat{\Gamma}\}$.  Regardless of whether or not $\vec{y}$ passes the test in
equation (\ref{eqn:midpt}), $\vec{y}$ is added to the set $\{\hat{Y}\}$ of all
simplex end points discovered by APS.

As a further guard against excessive simplex searching, all points $\vec{w}$ 
that pass the test in equation(\ref{eqn:simplexCriterion}) are further required
to satisfy the condition
\begin{eqnarray}
\chi^2(\vec{z})&\le&\chi^2(\vec{w})-0.25\times\left(\chi^2(\vec{w})-\chi^2(\vec{y})\right)\\
\vec{z}&\equiv&0.5\times\left(\vec{w}+\vec{y}\right)\\
\vec{y}&\in&\{\hat{Y}\}
\end{eqnarray}
before it is allowed to seed a new simplex search.  In this way, our APS avoids seeding simplex
searches with points that are inside of already known regions of low $\chi^2$.

@@

We make code implementing our extended APS for 
an arbitrary $\vec{\theta}\rightarrow\chi^2$
function available at \verb|https://github.com/uwssg/APS/|.
The code is written in C++.  It provides the option for several
user-specified parameters.  We list them below along with the C++ 
methods (belonging to the class \verb|aps|) used to set them.  
This code includes several user-specified parameters which we
outline below.

>>>>

We consider the following problem.  A researcher wants to constrain a theory
described by some number $N_p$ of tunable parameters, which we will represent by
the $N_p$-dimensional vector $\vec{\theta}$.  The available data measure some
function $f$ which is an implicit function of $\vec{\theta}$.  The researcher
is able to calculate a $\chi^2$ value quantifying the fit of a 
model $f(\vec{\theta})$ to the data and wishes to find either a confidence limit
defined by some rule, e.g $\chi^2\le\chi^2_\text{lim}$.  This is a general
phrasing of a Frequentist confidence limit as discussed in Appendix
\ref{sec:frequentism}.  We discuss alternatives for determining
$\chi^2_\text{lim}$ in Section \ref{sec:chi} below.  We will also consider
the case that the reseracher wishes to find Bayesian credible limt
to some probability $(1-\alpha)\%$.

@@

Markov Chain Monte Carlo (MCMC) methods propose to find the desired
limits by drawing random samples from the parameter space which are drawn
according to a probability density (the Bayesian posterior) which can be
interpreted as the probability that some value of $\vec{\theta}$ represents the
true value of $\vec{\theta}$ realized by the physical process underlying
the data.  
With enough of these samples, one can integrate this
probability density and find regions which contain some set amount of the total
probability.  In this case, ``credible limits'' are phrased as target values
for the probability integral $(1-\alpha)\%$, i.e. 
68\% for ``1-$\sigma$'' limits, 95\% for ``2-$\sigma$'' limits
and so on \cite{mcmc}.

We propose an alternative solution, using methods from machine learning
(to wit, Gaussian processes, though the algorithm admits many possible drivers; see
Section \ref{sec:gp})
to find the desired confidence limit by directly searching the
parameter space of $\vec{\theta}$ for points which lie on the boundary of the
desired confidence limit $\chi^2\le\chi^2_\text{lim}$.  
We refer to this algorithm as Active Parameter Searching (APS).
We present the algorithm in detail in Sections \ref{sec:algorithm} and
\ref{sec:gp}.
As it searches, APS uses
the knowledge it has already learned about $\vec{\theta}\rightarrow\chi^2$ 
to improve its search according to a metric that rewards the
algorithm both for finding points on the confidence limit and for sampling
points from previously unexplored regions of parameter space.  We find that this
behavior makes APS more robust against multi-modal $\chi^2$ functions (see
Section \ref{sec:toy}) without
sacrificing the speed of convergence 
associated with traditional MCMC methods (see Section \ref{sec:wmap}).
We show that, while APS is designed specifically to
yield Frequentist confidence
limits (which we define in Appendix \ref{sec:frequentism}),
it is also capable of roughly approximating
Bayesian credible limits on the space of
$\vec{\theta}$ (see Sections \ref{sec:bayes},\ref{sec:toy} and \ref{sec:wmap}).
This rough approximation may principally be of use in cases where little 
prior information about $\vec{\theta}$ exists and the researcher
requires APS's robustness against multi-modality to be certain that she
has discovered all of the best fits to her data.

@@

\begin{itemize}
\item(1A) Generate some initial number $N_s$ of randomly-distributed points in 
parameter space.  Find the values of $\chi^2$ corresponding to these points.
\\
\item(2A) Generate some number $N_c$ of candidate points.  
For each of these points, find the $N_g$ nearest points already sampled
by APS. Use these nearest neighbor points and their corresponding $\chi^2$ values
to predict the value of $\chi^2$ at each of these points.
 We will signify this prediction by $\mu$.  Calculate some value $\sigma$
 quantifying confidence in $\mu$ as a prediction of $\chi^2$.  Low values of
 $\sigma$ should correspond to confident predictions.
 \\
\item(3A) Choose the candidate point which maximizes the statistic
\begin{equation}
\label{eqn:sstat}
S=\sigma-|\mu-\chi^2_\text{lim}|
\end{equation}
and evaluate $\vec{\theta}\rightarrow\chi^2$ at that point.  Add this point and
its $\chi^2$ value to the list of sampled points begun in (1A). 
Return to (2A) and repeat until convergence.
\\
\end{itemize}
The confidence limit reported by APS is the list of all points found
for which $\chi^2\le\chi^2_\text{lim}$.  
These points, when plotted in one- or two-dimensional slices of parameter
space, ought to sketch out a region which contains the true value of $\vec{\theta}$
with the desired confidence (as defined in Appendix \ref{sec:frequentism}).
This is in contrast to the credible
limits of MCMC, which represent an integral over the sampled points containing some
fraction of the total posterior probability.  We present an alternative
interpretation for yielding credible limits from APS in Section \ref{sec:bayes}.

@@

The first possibility, a user-defined $\chi^2_\text{lim}$, is based
on the theory of Frequentist confidence intervals, which we briefly discuss
here and consider in much greater detail in Appendix \ref{sec:frequentism}.
If a data set is comprised of $N_d$ Gaussian-distributed data points
with a known covariance structure, then 
\begin{equation}
\chi^2\equiv\sum_{i,j}^{N_d}(d_i-f_i(\vec{\theta}))C^{-1}_{i,j}
(d_j-f_j(\vec{\theta}))
\end{equation}
will be distributed according to a $\chi^2$-distribution with $N_d$ degrees
of freedom.  This distribution can be integrated to determine the value
of $\chi^2$ containing $(1-\alpha)\%$ of the total probability.  This limiting
value will be the user-defined $\chi^2_\text{lim}$.

If the user does not feel comfortable settig $\chi^2_\text{lim}$
a priori, APS $\chi^2_\text{lim}$ can learn $\chi^2_\text{lim}$ by
defining it as $\chi^2_\text{min}+\Delta\chi^2$.  Here, $\chi^2_\text{min}$
is the minimum value of $\chi^2$ discovered by APS.  
$\Delta \chi^2$ is set
based on the properties of the $N_p$-dimensional parameter space being explored.
To wit, Wilks (1938) proves that $\Delta\chi^2$ for an $N_p$-dimensional
parameter space will be distributed according to a $\chi^2$-distribution
with $N_p$ degrees of freedom.  One can perform the same integral discussed
above and set $\Delta\chi^2$ to that value which encloses the desired confidence
limit.  This is the well-known Likelihood Ratio Test \cite{np}.
In this case, the user is relying on APS to eventually find the true
$\chi^2_\text{min}$ of the likelihood surface being explored.  
Section \ref{sec:simplex} below outlines an additional means of search
by which our expanded APS is capable of efficiently locating $\chi^2_\text{lim}$.

@@

Bryan {\it et al.} (2007) show that the APS search algorithm as presented in steps (1A-3A)
is robust against multi-modal
likelihood functions, using it to identify a disjoint region of high likelihood in
cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
spectrum.  This is the principal advantage to selecting points for evaluation according
to the statistic (\ref{eqn:sstat}).  Unfortunately, it is possible that the focus on 
exploring
unknown regions of parameter space might slow down convergence in the case of uni-modal
likelihood functions.  Therefore, we augment steps (1A-3A) with
the following modified search algorithms
to ensure that APS adequately characterizes known regions of high likelihood while
simultaneously searching for new regions to explore.

These modified searches will require APS to keep track of information it learns about the
$\vec{\theta}\rightarrow\chi^2$ fuction as it searches.  
To facilitate this discussion, we introduce the following notation.
Symbols denoted with $\vec{x}$ represent points in parameter space.  Variables denoted with
$\{\hat{X}\}$ denote sets of points in parameter space.  Variables denoted $\{X\}$ are sets
of numbers stored by APS.

@@

For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, APS performs the following:
\begin{itemize}
\item (1B) If fewer than $N_p$ focused searches have been proposed about $\vec{M}$, 
randomly choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
evaluate $\chi^2$ at that point.  
\\
\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
use bisection
along the line connecting the proposed point to $\vec{M}$ to find a point which does
satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line for which
$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
for which $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered in this way \
with $\chi^2=\chi^2_\text{lim}$ are stored in the set $\{\hat{B}(\vec{M})\}$ of 
``boundary points'' about $\vec{M}$ (each $\vec{M}$ in $\{\hat{\Gamma}\}$ will have its own
set $\{\hat{B}(\vec{M})\}$).  
\\
\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
Instead, examine the points in $\{\hat{B}(\vec{M})\}$ and identify the points
corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
points will be referred to as $\{\hat{B}_\text{MaxMin}(\vec{M})\}$.  
There will be $2\times N_p$ points in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, one for
the maximum and one for the minimum in each of our $N_p$ parameters.  
\\
\item(4B) For each point $\vec{b}$ in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, 
propose some number $N_v$ of new points (we have chosen
$N_v=20$ for the sake of speed) $\vec{v} = \vec{b}+\vec{\epsilon}$ where $\vec{\epsilon}$ is a
random, small vector in parameter space constrained to be perpendicular to
$\vec{b}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{v}$
($N_v$ such points for each of the $2\times N_p$ extremal boundary points in
$\{\hat{B}_\text{MaxMin}(\vec{M})\}$).
\\
\item(5B) At each point $\vec{v}$ above, find the $N_g$ nearest neighbor points as in step (2A) above
and evaluate equation (\ref{eqn:sstat}).  
Find the point $\vec{v}$ which maximizes $S$ and evaluate $\vec{\theta}\rightarrow\chi^2$ at
that point.  If
$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
Store the resulting $\chi^2=\chi^2_\text{lim}$ point in $\{\hat{B}(\vec{M})\}$
\end{itemize}

The modification suggested in section \ref{sec:focus} is designed to explore 
$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
perpendicularly to its ``radii'' (the vector $\vec{b}-\vec{M}$ would be a radius of
$\chi^2=\chi^2_\text{lim}$ described a sphere), with the hopes of being able
to extend the bounds of $\{\hat{B}_\text{MaxMin}(\vec{M})\}$.  In this way,
our extended APS attempts to plum the full Frequentist confidence limit $\chi^2\le\chi^2_\text{lim}$
as rapidly as possible.  However, we do not want this to come at the expense of the wide-ranging
exploration of steps (1A-3A).  Our version of APS alternates between performing steps (1A-3A) and steps
(1B-5B).  If there are $N_m$ distinct known local minima in $\{\hat{\Gamma}\}$, 
our algoritm will perform
$N_m$ iterations of steps (1A-3A), and then perform steps (1B-5B) once for each
$\vec{M}$ in $\{\hat{\Gamma}\}$.

@@

It is, however, also possible
that the $S$-maximization in steps (1A-3A) could also learn something about the
shape of the $\chi^2=\chi^2_\text{lim}$ contour, even if it does not place
points directly on that contour.  To explore that possibility, our
extended APS performs the
following modified search after each iteration of steps (1A-3A).

\begin{itemize}
\item(1C) For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, store, 
in addition to the boundary
points $\{\hat{B}(\vec{M})\}$, the projection of those boundary points
onto a unit sphere in parameter space, i.e. for each point $\vec{b}$ in $\{\hat{B}(\vec{M})\}$, draw a line
connecting $\vec{b}$ with $\vec{M}$ and keep track of
where a unit sphere centered on $\vec{M}$ intersects that line.  Call this set
of projections $\{\hat{U}\}$.
\\
\item(2C) Call the point for which $\chi^2$ has been evaluated in step (3A)
$\vec{P}$.  Find the
point $\vec{M}$ in $\{\hat{\Gamma}\}$ that is nearest to $\vec{P}$ in parameter space.
\\
\item(3C) Find the
projection of $\vec{P}$ onto the unit sphere surrounding $\vec{M}$.  
Call this projected point $\vec{p}$.  Find the point $\vec{u}$ in
$\{\hat{U}\}$ that is nearest to $\vec{p}$.  Record the parameter space 
distance between $\vec{u}$ and $\vec{p}$.
Add this distance to the set $\{D\}$ ($\{D\}$ will be a running list
of the $\vec{u}-\vec{p}$ distance for all points $\vec{P}$ found by steps
1A-3A, regardless of which point in $\vec{M}$ they are nearest).
\\
\item(4C) If the $\vec{p}-\vec{u}$ distance is greater than the 2/3 quantile of $\{D\}$,
then perform bisection between $\vec{M}$ and the point evaluated in step (3A).
Add the resulting $\chi^2=\chi^2_\text{lim}$ point to $\{\hat{B}(\vec{M})\}$.
\end{itemize}

Step (4C) ensures that, if the point $\vec{P}$ is in a novel direction from
$\vec{M}$, APS will find a point on the $\chi^2=\chi^2_\text{lim}$ contour
in that direction.  Because $\vec{P}$ was
chosen by maximizing equation (\ref{eqn:sstat}), we assume that there was something
``interesting'' about that direction from $\vec{M}$ (perhaps a sharp corner or distended
extension of the $\chi^2=\chi^2_\text{lim}$ contour).  In this way, our extended APS
will attempt to cover the surface of $\chi^2=\chi^2_\text{lim}$ in the most informative
way possible.

Because we expect the unit spheres surrounding the points $\vec{M}$ to become more densely
populated as the algorithm runs, we periodically delete the list $\{D\}$ so that
new directions away from $\vec{M}$ can still be explored.

@@

The final modification to the original APS algorithm which we propose in this
work is to utilize function minimization to identify local minima in $\chi^2$.
One of the principal drawbacks of the algorithm presented in steps (1A-3A) is
that it makes no account of ``near misses.''  If step (3A) finds that $\chi^2$
is lower than average but still greater than $\chi^2_\text{lim}$, the original
APS algorithm does nothing with that information.  We have tried to improve upon
that with our incorporation of bisection in sections \ref{sec:focus} and
\ref{sec:unitSphere} above.  Our final innovation is to allow APS to use points
with 
\begin{eqnarray}
\chi^2_\text{lim}&\le&\chi^2\le\chi^2_\text{threshold}\label{eqn:simplexCriterion}\\
\chi_\text{threshold}&\equiv&\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})
\label{eqn:threshold}
\end{eqnarray}
to seed a simplex function minimizer as developed by Nelder and Mead (1965).
Here, $G$ is a user-set parameter, $\chi^2_\text{min}$ is the minimum discovered
$\chi^2$ value and $\chi^2_{1/10}$ is the 1/10 quantile of the $\chi^2$ values
discovered by steps (1A-3A) (not those discovered by any of the modifications
discussed in sections \ref{sec:focus}, \ref{sec:unitSphere} or
\ref{sec:simplex}).

Because the Nelder-Mead simplex minimizer can be very time intensive, we balance our
algorithm by keeping track of how many times $\vec{\theta}\rightarrow\chi^2$
is called by the different searches and
demanding that the number of $\chi^2$ evaluations devoted to all
other variants of search (i.e. steps (1A-3A), (1B-5B), and (1C-4C)) equal the number of
$\chi^2$ evaluations devoted to the simplex minimizer before a new simplex
minimization can occur.

If the simplex minimizer converges to a point such that
$\chi^2<\chi^2_\text{lim}$, our extended APS determines whether or not it
has found a new local minimum to be added to $\{\hat{\Gamma}\}$ by comparing the end
point $\vec{y}$ of the minimization to each of the points already in
$\{\hat{\Gamma}\}$.  For each $\vec{M}$ in $\{\hat{\Gamma}\}$, APS calculates the mid-point
\begin{equation}
\label{eqn:midpt}
\vec{m}(\vec{M})=0.5\times\left(\vec{y}+\vec{M}\right)
\end{equation}
If, for all $\vec{M}$, 
$\chi^2$ at $\vec{m}(\vec{M})$ is greater
than $\chi^2_\text{lim}$, then it is deemed that $\vec{y}$ is a new local
minimum and is added to $\{\hat{\Gamma}\}$.  If not, $\vec{y}$ is assumed to be a
part of the one of the previously identified loci of low $\chi^2$ and it is not
added to $\{\hat{\Gamma}\}$.  Regardless of whether or not $\vec{y}$ passes the test in
equation (\ref{eqn:midpt}), $\vec{y}$ is added to the set $\{\hat{Y}\}$ of all
simplex end points discovered by APS.

As a further guard against excessive simplex searching, all points $\vec{w}$ 
that pass the test in equation (\ref{eqn:simplexCriterion}) are further required
to satisfy the condition
\begin{eqnarray}
\chi^2(\vec{z})&>&\chi^2(\vec{w})-0.25\times\left(\chi^2(\vec{w})-\chi^2(\vec{y})\right)\\
\vec{z}&\equiv&0.5\times\left(\vec{w}+\vec{y}\right)\\
\vec{y}&\in&\{\hat{Y}\}
\end{eqnarray}
before they are allowed to seed a new simplex search.  In this way, our 
extended APS avoids seeding simplex
searches with points that are inside of already known regions of low $\chi^2$.

@@

We make code implementing our extended APS for 
an arbitrary $\vec{\theta}\rightarrow\chi^2$
function available at \verb|https://github.com/uwssg/APS/|.
The code is written in C++.  It is designed for users to use
with any $\vec{theta}\rightarrow\chi^2$ function.  Users need only
to write a sub-class of the \verb|chisquared| class defined in
\verb|chisq.h| and \verb|chisq.cpp| which evaluates their desired
$\vec{\theta}\rightarrow\chi^2$ and pass an instantiation of this
sub-class to an instantiation of the \verb|aps| class defined in
\verb|aps.h| and \verb|aps.cpp|.  It is this \verb|aps| class which
does the work of performing the searches described above.
The user should not need to modify this class (though she is 
certainly welcome to).

Our \verb|aps| class provides the option for several
user-specified parameters.  We list them below along with the C++ 
methods (belonging to the class \verb|aps|) used to set them.  
This code includes several user-specified parameters which we
outline below.

\begin{itemize}
\item The parameter $G$ from equation (\ref{eqn:threshold}) above can be set using the method
\verb|aps.set_grat|
\\
\item $\chi^2_\text{lim}$ can be manually set using the method \verb|aps.set_target|
\\
\item $\Delta \chi^2$ can be set in the constructor for the class \verb|aps|.
\\
\item The number $N_g$ of nearest neighbors used in step 2A can be set in the constructor
for the class \verb|aps|
\\
\item $N_c$ from step 2A can be set using the method \verb|aps.set_samples|
\\
\item The minimum and maximum bounds in parameter space can be set using the 
method \verb|aps.set_max| and \verb|aps.set_min|.  These bounds will constrain the range of
points sampled in step 2A.  They will not contraint searches performed with bisection or the 
simplex minimizer.
\\
\item If there is a characteristic length scale associated with a parameter, this can be set
using the method \verb|aps.set_characteristic_length|.  This length scale will be used to
normalize all distances in parameter space.  If it is not set, it will default to the
difference between the maximum and minimum values set by \verb|aps.set_max| and
\verb|aps.set_min|.  Setting a characteristic length scale will principally
effect what points are considered nearest neighbors for the purposes of step 2A.
If the characteristic length scale of a parameter is large, that parameter will
be down-weighted when calculating the parameter-space distance between two points.
If the characteristic length scale of a parameter is small, that parameter will
be up-weighted in length calculations.
\end{itemize}
